---
title: "Prediction of Heart Disease"
author: "Claudia Lo" "Lindsay Maher"
date: "11/18/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

##Slide 1
Prediction of Heart Disease

##Slide 2
-H0: There is no relationship between the heart laboratory variables and heart disease, cardiovascular events or heart health
-Ha: Laboratory values have an effect on heart disease, cardiovascular events, and heart health

## R Markdown
```{r}
library(corrplot)
library(ggplot2)
library(polycor)
library(car)
library(pastecs)
library(psych)
library(ggm)
library(boot)
library(Hmisc)
library(plotly)
library(ggbeeswarm)
library(dplyr)
```
```{r}
getwd()
#setwd("C:/Users/claud/OneDrive/Desktop/fall 2021/stat comp/heart repo/prediction_of_heart_desease-project-4")
```



```{r}
#Import dataset
heart_disease<-read.csv("./data/heart.csv",header = TRUE, sep = ",")
```


```{r}
#check for NA's
any(is.na(heart_disease))
```


```{r}
summary(heart_disease)
str(heart_disease)
```


#Perform univariate distribution analysis for as many variables as possible, confirmed by quantitative values that describe the plots
```{r}
#histograms for numerical variables
hist(heart_disease$Age)
hist(heart_disease$RestingBP)
hist(heart_disease$Cholesterol)
hist(heart_disease$MaxHR)
hist(heart_disease$Oldpeak)
```


```{r}
heart_disease$Sex<-as.factor(heart_disease$Sex)
heart_disease$ChestPainType<-as.factor(heart_disease$ChestPainType)
heart_disease$FastingBS<-as.factor(heart_disease$FastingBS)
heart_disease$RestingECG<-as.factor(heart_disease$RestingECG)
heart_disease$ExerciseAngina<-as.factor(heart_disease$ExerciseAngina)
heart_disease$ST_Slope<-as.factor(heart_disease$ST_Slope)
str(heart_disease)
```

#visualizing our categorical variables 
```{r}
#barplot for Sex variable
df <- heart_disease %>%
		  filter(!is.na(Sex)) %>%  
		  group_by(Sex) %>%
  summarise(count_m= n())
		p <-ggplot(df, aes(x = Sex, y = count_m)) +
		  geom_bar(fill = "#B006F9", stat = "identity") +
		  geom_text(aes(label = count_m), vjust = -0.25,size = 4) + 
		  theme(axis.text = element_text(colour = "blue"),text = element_text(size=10),axis.line = element_line(size = 4, colour = "grey80"),axis.text.x = element_text(angle = 90, vjust = 0.45, hjust=1)) + ylab("Number of females and males in Sex category") + xlab("Sex") + ggtitle(labs(title = "Distribution of Sex"))
p
```



```{r}
#barplot for ChestPainType variable
df <- heart_disease %>%
		  filter(!is.na(ChestPainType)) %>%  
		  group_by(ChestPainType) %>%
  summarise(count_m= n())
		p <-ggplot(df, aes(x = ChestPainType, y = count_m)) +
		  geom_bar(fill = "#F9062B", stat = "identity") +
		  geom_text(aes(label = count_m), vjust = -0.25,size = 4) + 
		  theme(axis.text = element_text(colour = "blue"),text = element_text(size=10),axis.line = element_line(size = 4, colour = "grey80"),axis.text.x = element_text(angle = 90, vjust = 0.45, hjust=1)) + ylab("Number of types of chest pain") + xlab("Chest Pain Type") + ggtitle(labs(title = "Distribution of Chest Pain Types"))
p
```



```{r}
#barplot for FastingBS variable
df <- heart_disease %>%
		  filter(!is.na(FastingBS)) %>% 
		  group_by(FastingBS) %>%
  summarise(count_m= n())
		p <-ggplot(df, aes(x = FastingBS, y = count_m)) +
		  geom_bar(fill = "#3DDAEB", stat = "identity") +
		  geom_text(aes(label = count_m), vjust = -0.25,size = 4) + 
		  theme(axis.text = element_text(colour = "blue"),text = element_text(size=10),axis.line = element_line(size = 4, colour = "grey80"),axis.text.x = element_text(angle = 90, vjust = 0.45, hjust=1)) + ylab("Number of fasting blood sugars") + xlab("Fasting Blood Sugar") + ggtitle(labs(title = "Distribution of Fasting Blood Sugar"))
p
```

```{r}
#barplot for RestingECG variable
df <- heart_disease %>%
		  filter(!is.na(RestingECG)) %>%  
		  group_by(RestingECG) %>%
  summarise(count_m= n())
		p <-ggplot(df, aes(x = RestingECG, y = count_m)) +
		  geom_bar(fill = "#3DEB90", stat = "identity") +
		  geom_text(aes(label = count_m), vjust = -0.25,size = 4) + 
		  theme(axis.text = element_text(colour = "blue"),text = element_text(size=10),axis.line = element_line(size = 4, colour = "grey80"),axis.text.x = element_text(angle = 90, vjust = 0.45, hjust=1)) + ylab("Number of types of Resting ECG Results") + xlab("Resting ECG Results") + ggtitle(labs(title = "Distribution of Resting ECG Results"))
p
```



```{r}
#barplot for ExerciseAngina variable
df <- heart_disease %>%
		  filter(!is.na(ExerciseAngina)) %>%  
		  group_by(ExerciseAngina) %>%
  summarise(count_m= n())
		p <-ggplot(df, aes(x = ExerciseAngina, y = count_m)) +
		  geom_bar(fill = "#EB3DDB", stat = "identity") +
		  geom_text(aes(label = count_m), vjust = -0.25,size = 4) + 
		  theme(axis.text = element_text(colour = "blue"),text = element_text(size=10),axis.line = element_line(size = 4, colour = "grey80"),axis.text.x = element_text(angle = 90, vjust = 0.45, hjust=1)) + ylab("Number of females and males in Sex category") + xlab("Exercise Induced Angina") + ggtitle(labs(title = "Distribution of Exercise Induced Angina"))
p
```



```{r}
#barplot for ST_Slope variable
df <- heart_disease %>%
		  filter(!is.na(ST_Slope)) %>% 
		  group_by(ST_Slope) %>%
  summarise(count_m= n())
		p <-ggplot(df, aes(x = ST_Slope, y = count_m)) +
		  geom_bar(fill = "#EBA13D", stat = "identity") +
		  geom_text(aes(label = count_m), vjust = -0.25,size = 4) + 
		  theme(axis.text = element_text(colour = "blue"),text = element_text(size=10),axis.line = element_line(size = 4, colour = "grey80"),axis.text.x = element_text(angle = 90, vjust = 0.45, hjust=1)) + ylab("Number of types of ST segment slopes") + xlab("ST segment slopes") + ggtitle(labs(title = "Distribution of ST Segment Slopes"))
p
```

```{r}
df <- heart_disease %>%
		  filter(!is.na(FastingBS)) %>% 
		  group_by(HeartDisease) %>%
  summarise(count_m= n())
		p <-ggplot(df, aes(x = HeartDisease, y = count_m)) +
		  geom_bar(fill = "#3DDAEB", stat = "identity") +
		  geom_text(aes(label = count_m), vjust = -0.25,size = 4) + 
		  theme(axis.text = element_text(colour = "blue"),text = element_text(size=10),axis.line = element_line(size = 4, colour = "grey80"),axis.text.x = element_text(angle = 90, vjust = 0.45, hjust=1)) + ylab("positive for heart disease") + xlab("Heart Disease") + ggtitle(labs(title = "Distribution of Heart Disease"))
p
```

```{r}
#Document all the abnormalities observed and the decisions made to resolve them  during the exploratory data analysis
shapiro.test(heart_disease$Age)
shapiro.test(heart_disease$RestingBP)
shapiro.test(heart_disease$Cholesterol)
shapiro.test(heart_disease$MaxHR)
shapiro.test(heart_disease$Oldpeak)
stat.desc(cbind(heart_disease$Age,heart_disease$RestingBP,heart_disease$Cholesterol,heart_disease$MaxHR,heart_disease$Oldpeak), basic = FALSE, norm = TRUE)
#Since the p-values for Age, RestingBP,Cholesterol, MaxHR and Oldpeak are less than 0.05, we can reject H0 and say that the distribution is not normal.
```



#Plot grouped density and grouped histogram, to visualize frequency distributionof outcome variables and other variables in the dataset

```{r}
#RestingBP and Age_cat
heart_disease$Age_cat[heart_disease$Age>= 18 & heart_disease$Age<= 35]= "young adult"
heart_disease$Age_cat[heart_disease$Age>= 36 & heart_disease$Age<= 55]= "middle adult"
heart_disease$Age_cat[heart_disease$Age>= 56 & heart_disease$Age<= 65]= "older adult"
heart_disease$Age_cat[heart_disease$Age>= 66 & heart_disease$Age<= 110]= "elderly adult"

ggplot(heart_disease, aes(x = RestingBP, fill = Age_cat)) + geom_histogram(colour = "black",binwidth = .1, lwd = .75, linetype= 1,position = "identity", alpha = 4) + labs(x = "Resting BP", y = "Frequency", title = "Distribution of Resting BP by Age")

Densityplot <- ggplot(data = heart_disease,aes(x=RestingBP,fill= Age_cat ))+ geom_density(alpha = 0.9,cex=1)+ggtitle(labs(title = "Distribution of Resting BP by Age"))+theme(text = element_text(size=10))
Densityplot
```

```{r}
#MaxHR and Age_cat
heart_disease$Age_cat[heart_disease$Age>= 18 & heart_disease$Age<= 35]= "young adult"
heart_disease$Age_cat[heart_disease$Age>= 36 & heart_disease$Age<= 55]= "middle adult"
heart_disease$Age_cat[heart_disease$Age>= 56 & heart_disease$Age<= 65]= "older adult"
heart_disease$Age_cat[heart_disease$Age>= 66 & heart_disease$Age<= 110]= "elderly adult"

ggplot(heart_disease, aes(x = MaxHR, fill = Age_cat)) + geom_histogram(colour = "black",binwidth = .1, lwd = .75, linetype= 1,position = "identity", alpha = 4) + labs(x = "Max HR", y = "Frequency", title = "Distribution of Max HR by Age")

Densityplot <- ggplot(data = heart_disease,aes(x=MaxHR,fill= Age_cat ))+ geom_density(alpha = 0.9,cex=1)+ggtitle(labs(title = "Distribution of Max HR by Age"))+theme(text = element_text(size=10))
Densityplot
```
```{r}
summary(heart_disease$Cholesterol)
```

```{r}
#restingBP and Cholesterol_cat
heart_disease$Cholesterol_cat[heart_disease$Cholesterol>=0 & heart_disease$Cholesterol<= 200]= "normal"
heart_disease$Cholesterol_cat[heart_disease$Cholesterol>=201 & heart_disease$Cholesterol<= 603]= "high"

ggplot(heart_disease, aes(x = RestingBP, fill = Cholesterol_cat)) + geom_histogram(colour = "black",binwidth = .2, lwd = .75, linetype= 1,position = "identity", alpha = 4) + labs(x = "Resting BP", y = "Frequency", title = "Distribution of Resting BP by Cholesterol")

Densityplot <- ggplot(data = heart_disease,aes(x=RestingBP,fill= Cholesterol_cat ))+ geom_density(alpha = 0.7,cex=1)+ggtitle(labs(title = "Distribution of Resting BP by Cholesterol"))+theme(text = element_text(size=10))
Densityplot
```

```{r}
summary(heart_disease$RestingBP)
```

```{r setup, include=FALSE}
#MaxHR and RestingBP_cat
heart_disease$RestingBP_cat[heart_disease$RestingBP>= 0 & heart_disease$RestingBP<= 120]= "normal"
heart_disease$RestingBP_cat[heart_disease$RestingBP>= 121& heart_disease$RestingBP<= 200]= "high"


ggplot(heart_disease, aes(x = MaxHR, fill = RestingBP_cat)) + geom_histogram(colour = "black", binwidth = .2, lwd = .75, linetype= 1,position = "identity", alpha = 4) + labs(x = "Max Heart Rate", y = "Frequency", title = "Distribution of Max HR by RestingBP")

Densityplot <- ggplot(data = heart_disease,aes(x=MaxHR,fill= RestingBP_cat ))+ geom_density(alpha = 0.7,cex=1)+ggtitle(labs(title = "Distribution of Max HR by Resting BP"))+theme(text = element_text(size=10))
Densityplot

```

```{r}
heart_disease$HeartDisease<-as.factor(heart_disease$HeartDisease)
```

#Plot grouped categorical plots to visualize patterns that may exist in any relationships within the dataset

```{r}
#grouped barplot for HeartDisease and Age_cat
d4 <- heart_disease %>%
  group_by(HeartDisease,Age_cat) %>%
  summarise(count=n()) %>%
  mutate(perc=count/sum(count))
bar4 <- ggplot(data=d4, aes(x=HeartDisease, y=perc, fill=Age_cat)) +
  geom_bar(position = 'dodge', stat ='identity')+scale_y_continuous(labels = scales::percent)+labs(x = "Heart Disease", y = "Percentage of Heart Disease", fill = "Age_cat") + ggtitle(labs(title = "Heart Disease by Age"))+ theme(text = element_text(size=10))+ geom_text(aes(label = scales::percent(perc),y = perc),position=position_dodge(width=0.7), vjust = -0.25,size=3.5)
bar4  + facet_grid(. ~ HeartDisease,scales='free')
```

```{r}
#grouped bar plot for HeartDisease and restingBP_cat
d5 <- heart_disease %>%
  group_by(HeartDisease,RestingBP_cat) %>%
  summarise(count=n()) %>%
  mutate(perc=count/sum(count))
bar5 <- ggplot(data=d5, aes(x=HeartDisease, y=perc, fill=RestingBP_cat)) +
  geom_bar(position = 'dodge', stat ='identity')+scale_y_continuous(labels = scales::percent)+labs(x = "Heart Disease", y = "Percentage of Heart Disease", fill = "RestingBP_cat") + ggtitle(labs(title = "Heart Disease by Resting Blood Pressure"))+ theme(text = element_text(size=10))+ geom_text(aes(label = scales::percent(perc),y = perc),position=position_dodge(width=0.7), vjust = -0.25,size=3.5)
bar5  + facet_grid(. ~ HeartDisease,scales='free')
	
```


```{r}
#grouped bar plot for HeartDisease and Cholesterol_cat
d6 <- heart_disease %>%
  group_by(HeartDisease,Cholesterol_cat) %>%
  summarise(count=n()) %>%
  mutate(perc=count/sum(count))
bar6 <- ggplot(data=d6, aes(x=HeartDisease, y=perc, fill=Cholesterol_cat)) +
  geom_bar(position = 'dodge', stat ='identity')+scale_y_continuous(labels = scales::percent)+labs(x = "Heart Disease", y = "Percentage of Heart Disease", fill = "Cholesterol_cat") + ggtitle(labs(title = "Heart Disease by Cholesterol Level"))+ theme(text = element_text(size=10))+ geom_text(aes(label = scales::percent(perc),y = perc),position=position_dodge(width=0.7), vjust = -0.25,size=3.5)
bar6  + facet_grid(. ~ HeartDisease,scales='free')

```


```{r}
summary(heart_disease$MaxHR)
summary(heart_disease$Age)
```

```{r}
#grouped barplot for Age_cat and MaxHR_cat
heart_disease$MaxHR_cat[heart_disease$MaxHR>= 143 & heart_disease$MaxHR<= 192]= "normal"
heart_disease$MaxHR_cat[heart_disease$MaxHR<= 143]= "abnormal_1"
heart_disease$MaxHR_cat[heart_disease$MaxHR>= 192]= "abnormal_2"
table(heart_disease$MaxHR_cat)
```


```{r}
d6 <- heart_disease %>%
  group_by(Age_cat,MaxHR_cat) %>%
  summarise(count=n()) %>%
  mutate(perc=count/sum(count))
bar6 <- ggplot(data=d6, aes(x=Age_cat, y=perc, fill=MaxHR_cat)) +
  geom_bar(position = 'dodge', stat ='identity')+scale_y_continuous(labels = scales::percent)+labs(x = "Age_cat", y = "Max Heart Rate", fill = "MaxHR_cat") + ggtitle(labs(title = "Distribution of Age_cat and MaxHR_cat"))+ theme(text = element_text(size=10))+ geom_text(aes(label = scales::percent(perc),y = perc),position=position_dodge(width=0.7), vjust = -0.25,size=3.5)
bar6  + facet_grid(. ~ Age_cat,scales='free')
```

#Build correlation matrix to confirm the magnitude of the correlation coefficient and direction of observed relationships
```{r}
heart_disease_sub_num<-heart_disease[,c("Age","RestingBP","Cholesterol", "MaxHR", "Oldpeak")]
```

```{r setup, include=FALSE}
#RestingBP and Age
cor(heart_disease_sub_num, use="complete.obs", method="pearson")
cor.test(heart_disease$RestingBP,heart_disease$Age,alternative="less",method="pearson")

c<-cor(heart_disease_sub_num,use="complete.obs", method="pearson")
corrplot(c,method="color",title = "Relationship between  variables in heart disease",addgrid.col ="grey",addCoef.col = "pink")
scatterplotMatrix (heart_disease_sub_num, spread=FALSE, smoother.args=list(lty=2), main="Scatter Plot Matrix")
```

#Use the observed correlation relationships to predict the outcome variable with some form of regression 
```{r}
#logistic regression

model.1<-glm(HeartDisease~Age+Sex+ChestPainType+RestingBP+Cholesterol+FastingBS+RestingECG+MaxHR+ExerciseAngina+Oldpeak+ST_Slope,data=heart_disease,family=binomial())

model.2<-glm(HeartDisease~Age+ChestPainType+RestingBP+Cholesterol+RestingECG+MaxHR+ExerciseAngina+Oldpeak,data=heart_disease, family=binomial())

model.3<-glm(HeartDisease~ChestPainType+RestingBP+ExerciseAngina,data=heart_disease,family=binomial())
```


```{r}
summary(model.1)
summary(model.2)
summary(model.3)
```

```{r}
logisticPseudoR2s <- function(LogModel) {
	dev <- LogModel$deviance 
	nullDev <- LogModel$null.deviance 
	modelN <- length(LogModel$fitted.values)
	R.l <-  1 -  dev / nullDev
	R.cs <- 1- exp ( -(nullDev - dev) / modelN)
	R.n <- R.cs / ( 1 - ( exp (-(nullDev / modelN))))
	cat("Pseudo R^2 for logistic regression\n")
	cat("Hosmer and Lemeshow R^2  ", round(R.l, 3), "\n")
	cat("Cox and Snell R^2        ", round(R.cs, 3), "\n")
	cat("Nagelkerke R^2           ", round(R.n, 3),    "\n")
      }
```

```{r}
logisticPseudoR2s(model.1)
logisticPseudoR2s(model.2)
logisticPseudoR2s(model.3)
```

```{r}
#compute predicted readmission status
heart_disease$predict <- predict(model.1, newdata = heart_disease, type = "response" )
heart_disease$predicted_outcome <- ifelse(heart_disease$predict > 0.5,1,0)
```


```{r}
table(heart_disease$HeartDisease)
```

```{r}
#build confusion matrix
heart_disease$conf_mat[heart_disease$predicted_outcome == 1 & heart_disease$HeartDisease == 1]= "True positive"
heart_disease$conf_mat[heart_disease$predicted_outcome == 1 & heart_disease$HeartDisease == 0]= "False positive"
heart_disease$conf_mat[heart_disease$predicted_outcome == 0 & heart_disease$HeartDisease == 1]= "False negative"
heart_disease$conf_mat[heart_disease$predicted_outcome == 0 & heart_disease$HeartDisease == 0]= "True negative"
```


```{r}
table(heart_disease$conf_mat)
```
```{r}
#accuracy
((458+341)/(458+341+69+50))
```


#Conduct bias analysis with a demographic variable if necessary
```{r}
#Sex
table(heart_disease['Sex'])
CI_normalize<-(193-725)/(193+725)
CI_normalize
```


